Formal Analysis of EDHOC Key Establishment for Constrained IoT Devices

### 2022-07-13 paper.tex
[add] more elaboration on DoS

\begin{document}
\maketitle
%
\begin{abstract}
Constrained IoT devices are becoming ubiquitous in society
and there is a need for secure communication protocols that respect the
constraints under which these devices operate.
%
\mEdhoc{} is an authenticated key establishment protocol for constrained IoT
devices, currently being standardized by the Internet Engineering Task
Force (IETF).
%
A rudimentary version of \mEdhoc{} with only two key establishment methods was
formally analyzed in 2018.
%
Since then, the protocol has evolved significantly and several new key
establishment methods have been added.
%
In this paper, we present a formal analysis of all \mEdhoc{} methods in an
enhanced symbolic Dolev-Yao model using the \mTamarin{} tool.
%
We show that not all methods satisfy the authentication notion
injective agreement, but that all do satisfy a notion of implicit
authentication, as well as Perfect Forward Secrecy (PFS) of the session key material.
%
We identify other weaknesses to which we propose improvements
%
For example, a party may intend to establish a session key with a certain peer,
but ending up establishing it with another, trusted but compromised, peer.
%
We communicated our findings and proposals to the IETF, which
has incorporated some of these in newer versions of the standard.
\end{abstract}
%
%-------------------------------------------------------------------------- sec
\section{\uppercase{Introduction}}
\label{sec:introduction}

As IoT devices become more prevalent and get involved in progressively sensitive
functions in society, the need to secure their communications
becomes increasingly important.
%
Most security analysis has focused on computationally
strong devices, such as cars and web-cameras, where existing protocols like
\mDandTls{} suffice.
%
Constrained devices, on the other hand, which operate under severe
bandwidth and energy consumption restrictions, have received much less
attention.
%
These devices may be simple sensors, which only relay environment
measurements to a server every hour, but need to function autonomously without
maintenance for long periods of time.
%
The IETF standardized the Object Security for
Constrained RESTful Environments (\mOscore{}) protocol to secure communications
between constrained devices~\cite{rfc8613}.
%
However, the \mOscore{} protocol requires a pre-established security context.
%
The IETF has been discussing requirements and mechanisms for a key
exchange protocol, named Ephemeral Diffie-Hellman Over COSE (\mEdhoc), for
establishing \mOscore{} security contexts.
%
Naturally, \mEdhoc{} must work under the same constrained requirements as
\mOscore{} itself.
%
While not all use cases for \mEdhoc{} are firmly set, the overall goal is to
establish an \mOscore{} security context, under message size limitations.
%
It is therefore important to ensure that \mEdhoc{} satisfies fundamental
security properties expected from a key exchange protocol.
%The work is now being done in the IETF Lightweight Authenticated Key Exchange
%(LAKE) working group.
%

The first incarnation of \mEdhoc{} appeared in March 2016.
%
It contained two different key establishment methods, one based on a
pre-shared Diffie-Hellman (DH) cryptographic core%
\footnote{By a \emph{cryptographic core}, or simply core, we mean an academic protocol,
without encodings or application specific details required by an industrial
protocol.
%
A \emph{key establishment method} is a core with some such details added.
}
and a second based on a
variant of challenge-response signatures in the style of
\mOptls{}~\cite{DBLP:conf/eurosp/KrawczykW16}.
%
\mEdhoc{} is therefore a framework of several key establishment methods.
%
In May 2018, the core based on challenge-response signatures was replaced by
one based on \mSigma{} (SIGn-and-MAc)~\cite{sigma,bruni-analysis-selander-ace-cose-ecdhe-08}.
%
Since then the protocol has undergone significant changes.
%
Three new cores, mixing challenge-response signatures and regular signature for
authentication, were added~\cite{our-analysis-selander-lake-edhoc-00}.
%

We formulate and formalize a security model covering all four key
establishment methods, which is important especially since the
\mSpec{}~\cite{our-analysis-selander-lake-edhoc-00} lacks a clear description
of the intended security model and overall security goals.
%

We perform the analysis in a symbolic Dolev-Yao model.
In this framework, we model messages as terms in an algebra, with operations such
as encryption, modelled as functions on these terms.
These functions are assumed perfect, e.g., one cannot decrypt an encrypted
message without access to the key.
The adversary, while unable to break encryption or reverse hashing, is modelled
as the network.
That is, the adversary, can block reroute, replay and modify messages at will.
A symbolic model like this, while slightly severe an abstraction, still allows
us to analyze \mEdhoc{} for logical flaws without incurring the complexity of
a computational model.
%
The standardization process is ongoing, with the authors releasing newer
versions of the \mSpec{} (see Section~\ref{sec:newdrafts} for more detail
about how these versions differ from the one analyzed here).
%

%-------------------------------------------------------------------------- sub
\subsection{Contributions}
\label{sec:contributions}
In this paper, we formally analyze the \mEdhoc{} protocol (with its four key
establishment methods) using the \mTamarin{} tool~\cite{DBLP:conf/cav/MeierSCB13}.
%
We present a formal model we constructed of the protocol as given in the
\mSpec{}~\cite{our-analysis-selander-lake-edhoc-00}.
%

We give an explicit adversary model for the protocol and verify
properties such as session key material and entity authentication, and perfect
forward secrecy, for all four methods.
%

The model itself is valuable as a basis for verifying further updates in the
ongoing standardization.
%
It is publicly available~\cite{edhocTamarinRepo}.
%
It took several person-months to interpret the
specification and construct the model.
%
Termination requires a hand-crafted proof oracle to guide \mTamarin{}.
%

We show that not all \mEdhoc{}'s key establishment
methods provide authentication according to the injective agreement definition
on the session key material, and none on the initiator's identity.
%
However, we show that all methods fulfill an implicit agreement property
covering the session key material and the initiator's identity.
%
We identify a number of subtleties, ambiguities and weaknesses in the
specification.
%
For example, the authentication policy requirements allows situations where a
party establishes session key material with a trusted but compromised peer, even
though the intention was to establish it with a different trusted party.
%
We provide remedies for the identified issues and have
communicated these to the IETF and the specification authors, who incorporated
some of our suggestions and currently consider how to deal with the remaining
ones.
%

%-------------------------------------------------------------------------- sub
\subsection{Comparison with Related Work}
The May 2018 version of \mEdhoc{} was formally analyzed by
\cite{DBLP:conf/secsr/BruniJPS18} using the \mProverif{}
tool~\cite{DBLP:conf/csfw/Blanchet01}.
%
Their analysis covered a pre-shared key authenticated core and one
based on \mSigma.
%
The properties checked for therein were secrecy, PFS and integrity of
application data, identity protection against an active adversary,
and strong authentication.
%

In contrast to the key establishment methods analyzed by Bruni et~al., which
were based on the well-understood pre-shared key DH and \mSigma{} protocols,
the three newly added
methods combine two unilateral authentication protocols with the goal to
constructing mutual authentication protocols.
%
Combining two protocols, which individually provide unilateral authentication,
is not guaranteed to result in a secure mutual authentication
protocol~\cite{DBLP:conf/ccs/Krawczyk16}.
%
Consequently, even though the framework is similar to the one analyzed by Bruni
et~al., the cryptographic underpinnings have significantly increased in
complexity, and is using mechanisms which have not previously been formally analyzed.
%
The set of properties we check for is also different.
%
Our analysis is further carried out using a different tool,
namely \mTamarin; different kinds of strategies to formulate and
successfully analyze the protocol are required when working with this tool.
%

%-------------------------------------------------------------------------- sec
\section{\uppercase{The \mEdhoc{} Protocol}}
\label{sec:edhoc}
\input{edhocProtocol}

%-------------------------------------------------------------------------- sec
\section{\uppercase{Formalization and Results}}
\label{sec:formalization}
\input{edhocFormalization}

%-------------------------------------------------------------------------- sec
\section{\uppercase{Denial of Service}}
\label{sec:dos}
%\mPoint{DoS is out of scope for \mEdhoc{}; apps to deal with it}
\mEdhoc{} includes no specific measures for countering Denial of Service (DoS)
attacks and delegates the responsibility to handle this to the application.
%
The core \mEdhoc{} specification has limited functionality, few negotiable
parameters and processing rules.
%
Hence, we argue in this section that its attack surface
is limited from a DoS perspective.
%
However, \mEdhoc{} leaves much of the semantics of its messages undefined or
specified as (optional) extensions.
%
It is therefore difficult to draw the line between what is \mEdhoc{} and what is
an extension.
%
For example, the core specification does not include a
message transport mechanism.
%
This increases the risk of applications vulnerable to DoS via \mEdhoc{}, even
though \mEdhoc{} itself is arguably not particularly weak in this respect.
%

@@

No recommendation is given on how to minimize DoS effects stemming from use of
error messages, which can be used in an arbitrary way in \mEdhoc{}.
%
While there is some work on DDoS aspects of \mCoap{}, and much on IoT in
general, this does not capture the particular consequences originating with
\mEdhoc{}.
%
We now elaborate on some considerations application developers needs to take
into account.
%

@@

Depending on the use case, the first \mEdhoc{} message may be sent by either
party in a session.
%
For example, even though it may seem intuitive that an electronic
key-card initiates the connection towards lock, the first \mEdhoc{} message
could be included in the response from the lock.
%
This means the lock can postpone heavy verification actions until it has
authenticated the key card after receiving the third \mEdhoc{} message.
%

@@

\textbf{Q:} Is htere any situation where it makes sense that the initiator sends
a unsupported ciphersuite message instead of message 3?
%

@@

Section 6.2 gives an excample of sending an error message "method not
supported".
%
This message cannot be used to negotiate method securely.
%
The reason is that error messages are not integrity protected, and an initator
only proposes one method, and therefore the same tactic used for ciphersuite
negotiation cannot be used for methods.
%
Because method negotiation can be viewed in the context of crypto agility, there
is room for improvement.
%

>>>>

\begin{document}
\maketitle
%
\begin{abstract}
Given how common IoT devices that use constrained resources are becoming today, the need of the hour is communication protocols which can operate securely under such limitations.
%
For a few years, the Internet Engineering Task Force (IETF) has been working to standardize \mEdhoc{}, an authenticated key establishment protocol for such constrained IoT devices.
%
The first version of \mEdhoc{} was proposed in 2016.
%
In 2018, Bruni et al~\cite{DBLP:conf/secsr/BruniJPS18} used the \mProverif{} tool~\cite{DBLP:conf/csfw/Blanchet01} to formally analyze an early version of \mEdhoc{}, which had only two key establishment methods.  
%
%By now, the protocol has been fleshed out much more, and has been augmented with multiple new key establishment methods.
By 2021, the protocol had been fleshed out much more, with multiple new key establishment methods, and this version was formally analyzed using the \mTamarin{} prover~\cite{DBLP:conf/cav/MeierSCB13} in~\cite{Norr21}.
%
Here, we build on this work, and use \mTamarin{} to formally analyze the key establishment methods in the current version of \mEdhoc{}, as well as discuss some ramifications of the choices made while designing the protocol.
\vnote{We need to see what more we can say here, about the new stuff we're potentially adding}
\end{abstract}
%
%-------------------------------------------------------------------------- sec
\section{\uppercase{Introduction}}
\label{sec:introduction}
\input{intro}

%-------------------------------------------------------------------------- sec
\section{\uppercase{The \mEdhoc{} Protocol}}
\label{sec:edhoc}
\input{protocol}

%-------------------------------------------------------------------------- sec
\section{\uppercase{Formalization and Results}}
\label{sec:formalization}
\input{formalization}

%-------------------------------------------------------------------------- sec
\section{\uppercase{Denial of Service}}
\label{sec:dos}

% There is little about EDHOC specific DoS stuff; that is interesting in itself,
% and motivates this section
There is a wealth of information about DoS in general, and more specifically for
IoT.
%
However, there is little information regarding DoS aspects stemming
specifically from \mEdhoc{}.
%

% Other protocols have DoS considerations, but EDHOC delegates it to developer
In contrast to \mTls{} and IKEv2,
the \mEdhoc{} core specification does not include any specific measures for
countering Denial of Service (DoS) attacks.
%
Instead, that responsibility is delegated to the application, using
proprietary measures or extensions to \mEdhoc{}.
%

% The core is small and probably not vulnerable to anything unexpected
The \mEdhoc{} core specification has limited functionality, few negotiable
parameters and processing rules.
%
Hence, we argue in this section that its attack surface
is limited from a DoS perspective.
%

% However, It is difficult to define what EDHOC *is*; it s partially specified in many
% specs
However, \mEdhoc{} leaves much of the semantics of its messages undefined or
specified as extensions.
%
Examples of mechanisms with mostly undefined semantics include processing
rules and content of error messages and of the additional data that is
passed in each of the \mEdhoc{} messages.
%
Another example is that the core specification does not include a
message transport mechanism.
%
The transport for \mEdhoc's primary target, i.e., to key \mOscore{},
is \mCoap{}.
%
This is specified in~\cite{ietf-core-oscore-edhoc-03}, but still leaves
many design choices to the application.
%
Because of this, it is difficult to draw the line between what is considered as
\mEdhoc{} and what is an extension.
%

% Pros with undefined semantics: flexibilty; Cons: developers can shoot
% themselves in the foot
Undefined semantics allow the implementors much freedom
and flexibility to design applications.
%
On the down-side, the few recommendations given increase the risk of
applications being vulnerable to DoS via \mEdhoc{}, even
though the core \mEdhoc{} specification itself is arguably not particularly
weak in this respect.
%

@@

No recommendation is given on how to minimize DoS effects stemming from use of
error messages, which can be used in an arbitrary way in \mEdhoc{}
(see Section~\ref{sec:errorHandling} below).
%
While there is some work on DDoS aspects of \mCoap{}, and much on IoT in
general, this does not capture the particular consequences originating with
\mEdhoc{}.
%
We now elaborate on some considerations application developers needs to take
into account.
%

@@

Depending on the use case, the first \mEdhoc{} message may be sent by either
party in a session.
%
For example, even though it may seem intuitive that an electronic
key-card initiates the connection towards lock, the first \mEdhoc{} message
could be included in the response from the lock.
%
This means the lock can postpone heavy verification actions until it has
authenticated the key card after receiving the third \mEdhoc{} message.
%
The draft IETF specification profiling \mEdhoc{} for use with
\mCoap{} and \mOscore{} ~\cite{ietf-core-oscore-edhoc-03},
however, defines the
\mCoap{} client to be the \mEdhoc{} initiator, and hence reduces the flexibility
to use that option.
%

@@

\textbf{Q:} Is there any situation where it makes sense that the initiator sends
a unsupported ciphersuite message instead of message 3?
%

@@

Section 6.2 gives an example of sending an error message "method not
supported".
%
This message cannot be used to negotiate method securely.
%
The reason is that error messages are not integrity protected, and an initator
only proposes one method, and therefore the same tactic used for ciphersuite
negotiation cannot be used for methods.
%
Because method negotiation can be viewed in the context of crypto agility, there
is room for improvement.
%
